---
title: "Simonson_HW4"
author: "Marty Simonson"
date: "February 14, 2019"
output: pdf_document
editor_options: 
  chunk_output_type: console
---
```{r, include = F}
packages<-function(x, repos="http://cran.r-project.org", ...){
x<-as.character(match.call()[[2]])
if (!require(x,character.only=TRUE)){
install.packages(pkgs=x, repos=repos, ...)
require(x,character.only=TRUE)
}
}

packages(ggplot2)
```
# 1.) Pollen Removal

Using data from _bee.csv_, test if log-transformation helps meet assumptions of pooled t-tests.

a) Construct box plots and normal quantile plots on the log-transformed duration values for both types of bees. 
```{r}
df<-read.csv("bee.csv")
df$LogDuration<-log(df$Duration)

ggplot(data=df,aes(x=Type, y=LogDuration))+
       geom_boxplot(fill="aliceblue")+
       ylab("Log(Duration)")+
       theme_classic()

ggplot(data=df, aes(sample=LogDuration)) + 
        stat_qq() + 
        stat_qq_line() + 
        facet_wrap(facets = vars(Type))+
        theme_classic()
```

Does it appear that assuming normality for thelog-transformed duration variable is reasonable? Explain.
- __Answer:__ The assumption of normality is not met after log-transformation because there is still curvature in the points along the QQ line.

Are the variabilities the same for the two types of bees on the log scale? Explain.
- __Answer:__ The size of the boxes and whiskers are not equal, therefore the assumption of normality is not met.

b) Use the Welch's t-test to compare the (population) median durations. Write down the3 null and alternate hypotheses, provide the t-stastic and p-value.