---
title: "Lab 11"
author: "Marty Simonson"
date: "April 17, 2019"
output: html_document
editor_options: 
  chunk_output_type: console
---

Libraries

```{r}
library(ggplot2)
library(emmeans)
library(car)
```

EXCEPT in polynomial regression, it is always wise to use sum-contrasts for factors in stead of treatment contrast. This can be globally set by :

```{r}
options(contrasts=c(unordered="contr.sum",ordered="contr.poly"))
```


1.) 

```{r}
corn<-read.table("Data/ManureFert.dat",header=T)
str(corn)
```


find the interaction plot by fitting an interaction model

```{r}
fit.corn<-lm(Yield~Manure*Fertilizer, data = corn)
```

interaction plot from "emmip" from the emmeans package

```{r}
emmip(fit.corn, Manure ~ Fertilizer)+theme_classic()+ylab("Mean Yield")
```

We see graphical evidence for an interaction between manure and fertilizer. There does not seem to be an effect of fertilizer when manure is high.

The table of partial f-tests can be obtained using the "Anova" command with "type=3" specified. (ignore intercept, not explained why)

```{r}
Anova(fit.corn, type = 3)
```

The f-ratio is 1.8053 with 24 and 2 degrees of freedom. The p-value is 0.186 and provides very weak evidence for an interaction.

----

To answer a farmer's question we use all pairwise comparisons among the treatment means (simple ffects) and use a Tukey's adjustment for multiple comparisons. Because the treatments are simply the combinations of two factors, we can obtain them by:

```{r}
trt.emm<-emmeans(fit.corn, ~ Manure * Fertilizer)
trt.emm
# paired comparisons
pairs(trt.emm)
# connected letter diagram
CLD(trt.emm)
```

no evidence for a clear winner. the farmer can choose any of the treatmens except the control that fits their budget. 

In order to estimate and test for a manure effect in the presenece of ertilizer, we want ot check the contrast .5(high:present)+.5(low:present)-(none:present).



```{r}
# Note the ordering of the treatments
trt.emm
# the terms involved are the last three in this object

contrast(trt.emm, list("manure effect in presence of fert."=c(0,0,0,.5,.5,-1)))
# strong evidence that the use of manure increases yield in presence of fertilizer.
```

to estimate and test for a fertilizer effect at high level of manure

```{r}
contrast(trt.emm, list("fertilizer effect at high manure"=c(-1,0,0,1,0,0)))
#no evidence for an effect of fertilizer at high manure doses.
```

----

2: chromotagraphy
```{r}
chrom<-read.table("Data/chromatography.txt",header=T)
fit.int<-lm(logCounts~Concentration*FlowRate, data = chrom)
summary(fit.int)

# plot
emmip(fit.int, FlowRate ~ Concentration) + theme_classic()+ylab("mean log-count")

# Type 3 anova table
Anova(fit.int, type = 3)
# appears that here is no evidence of interaction
```

now without interaction

```{r}
fit<-lm(logCounts ~ Concentration + FlowRate, data = chrom)
Anova(fit)
# seems that there is strong evidence for each of the main effects
```

compare multiple comparisons to determine which concentrations havae signifciantly different mean(log) counts.

```{r}
conc.emm<-emmeans(fit,~Concentration)

conc.emm

pairs(conc.emm)

CLD(conc.emm, Letters = LETTERS)
# it appears that each concentration has different mean (log) counts
```




















