---
title: "Lab10A"
author: "Marty Simonson"
date: "April 10, 2019"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r}
library(ggplot2)
library(car)
```

Read cigarette data:

```{r}
cig<-read.table("Data/cigarette.dat",header=T)
head(cig)
str(cig)
```

scatterplot matrix:

```{r}
pairs(cig)
```

pairwise correlation coefficients:

```{r}
cor(cig)
```

Fit the Multiple Linear Regression model

```{r}
f0<-lm(CO~TAR + Nicotine + Weight, data = cig)
```

VIF by using the command "vif" from library "car"

```{r}
vif(f0) # Variance Inflation Factors for each variable
```

drop the "weight" variable

```{r}
f1<-lm(CO~TAR+Nicotine,data=cig)
summary(f1)
```

drop the "tar" variable

```{r}
f2<-lm(CO~Nicotine,data=cig)
summary(f2)
```

Replace nicotine with TAR

```{r}
f3<-lm(CO~TAR,data=cig)
summary(f3)
```

predict using the goldflake honeydew data

model 1

```{r}
goldflake<-data.frame(Weight=0.845,Nicotine = 0.7, TAR = 10.3)
predict(f1, newdata = goldflake, se.fit = T)
```

model 2

```{r}
predict(f2, newdata = goldflake, se.fit = T)
```

model 3

```{r}
predict(f3, newdata = goldflake, se.fit=T)
```

check the AICc for each of the three models:

model 1-3

```{r}
AIC(f1)
AIC(f2)
AIC(f3)
```

check the BIC

```{r}
BIC(f1)
BIC(f2)
BIC(f3)
```

#2: SAT scores: an observational study

Takers: percent of total eligible students in the state who took the exam;
Income: is the median income of families of test atkers (100s of dollars)
Years: number of years that the test-takers had formal studies in test topics
Public: percentage of test-takers whoa ttend public secondary schools
Expend: total state expenditure on secondary schools, (100s of dollars per student)
rank: median percentile raking of the test-takers within their secondary school classes

```{r}
sat<-read.csv("Data/sat.csv",header=T)
str(sat)
```

bare model: just percentage of takers and median class rank

```{r}
mod0<-lm(sat~takers+rank, data = sat)
summary(mod0)
```

first model: fully saturated

```{r}
mod1<-lm(sat~takers+income+years+public+expend+rank, data = sat)
summary(mod1)
```

mod2: no expenditures

```{r}
mod2<-lm(sat~takers+income+years+public+rank, data = sat)
summary(mod2)
```

mod3: no public

```{r}
mod3<-lm(sat~takers+income+years+expend+rank, data = sat)
summary(mod3)
```

mod4: no years

```{r}
mod4<-lm(sat~takers+income+public+expend+rank, data = sat)
summary(mod4)
```

mod5: no income

```{r}
mod5<-lm(sat~takers+public+years+expend+rank,data = sat)
summary(mod5)
```

mod6: no public or expenditures

```{r}
mod6<-lm(sat~takers+income+years+rank, data = sat)
summary(mod6)
```

mod7: no years or expenditures

```{r}
mod7<-lm(sat~takers+rank+income+public, data = sat)
summary(mod7)
```

mod8: no income or expenditures

```{r}
mod8<-lm(sat~takers+rank+years+public, data = sat)
summary(mod8)
```

mod9: no public or years

```{r}
mod9<-lm(sat~takers+rank+income+expend,data=sat)
summary(mod9)
```

mod10: no public or income

```{r}
mod10<-lm(sat~takers+rank+years+expend, data=sat)
summary(mod10)
```

mod11: no years or income

```{r}
mod11<-lm(sat~takers+rank+expend+public,data=sat)
summary(mod11)
```

```{r}
AIC(mod1)
AIC(mod2)
AIC(mod3)
AIC(mod4)
AIC(mod5)
AIC(mod6)
AIC(mod7)
AIC(mod8)
AIC(mod9)
AIC(mod10)
AIC(mod11)
```

model 7 is best AIC value

```{r}
BIC(mod1)
BIC(mod2)
BIC(mod3)
BIC(mod4)
BIC(mod5)
BIC(mod6)
BIC(mod7)
BIC(mod8)
BIC(mod9)
BIC(mod10)
BIC(mod11)
```

model 7 is best BIC value

```{r}
summary(mod7)
```

